The stopping power of hydrogen in hafnium and the importance of relativistic $4f$ electrons

### 2020-04-20 april_version/Stopping_Hf_abril20.tex
checked with grammarly

%Title of paper
\title{The stopping power of \clau{hydrogen in} hafnium and the importance of relativistic $4f$ electrons}
%Authors and affiliations

@@

%---------------------------Chile
\author{\clau{P. A. Miranda}}
\affiliation{Departamento de F\'isica, Facultad de Ciencias Naturales, 
Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 
7800002, Chile}
%-------------------------Lisboa
\author{\clau{E. Alves}}
\affiliation{Centro de Ci\^{e}ncias e Tecnologias Nucleares, Instituto 
Superior T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, 
Portugal}
\affiliation{Instituto de Plasma e Fus\~{a}o Nuclear, Instituto Superior 
T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, Portugal}

@@

So far, only one experimental work has been published regarding the 
stopping power cross section of pure hafnium for protons~\cite{Sirotinin}, 
while more attention has been recently given to  studies involving 
hafnium oxide due to its practical use~\ale{\cite{Abril,Behar,Primetzhofer,Roth}}. 
It is well known that significant attention has been paid in recent 
years to transition metal-oxides such as HfO$_2$ because of their 
potential as alternative gate dielectrics to replace SiO$_2$ for the 
future generation of nano-electronics with less than 45 nm gate 
length~\cite{Choi,Robertson}. Some important physical properties of the 
above mentioned metal-oxide films depend on its thickness, which is 
often measured by using Rutherford Backscattering Spectrometry~\cite{Alfassi01,Tesmer01}, 
a method that relies heavily on the determination of the stopping power 
of ion beams in the material of interest.

@@

The theoretical approach implemented in this work uses the shell-wise local plasma approximation (SLPA)~\cite{mon13} to describe the energy 
transferred to the bound $1s$-$4f$ electrons, and two different models 
for the FEG: \clau{in low energy region,}
the screened potential with cusp condition model (SPCC)~\cite{mon17}, 
which is non-linear binary formalism; and the Mermin-Lindhard dielectric 
formalism (ML)~\cite{Mermin}, for energies around the stopping maximum 
and above. Our model requires the relativistic wave functions and 
binding energies of Hf, and considers 4 electrons per atom in the 
FEG~\cite{mendez2019}. Hf has the extra interest of the filled 
$4f$-subshell (with 14 electrons) as the main contribution below the 
FEG, causing the stopping cross sections to be very sensitive to a good 
description of this shell. The screening among the $4f$ and $5p$ 
electrons has been considered and found to play a major role within the 
SLPA calculations.

The experimental details and data are given in Section~\ref{experiment}, 
while the theoretical method is explained in Section~\ref{theory}. 
Present theoretical and experimental values are finally compared to the \clau{only experimental values by Sirotinin group~\cite{Sirotinin} measured with backscattering method, the theoretical results by  Grande and Schiwietz~\cite{Grande,casp52}, and by Sigmund and Schinner~\cite{DPASS20}, and also with the semi-empirical values from the SRIM-2013 package~\cite{Ziegler01} and the ICRU-49 tabulation~\cite{ICRU49}}. 
Conclusions and discussions are given in Section~\ref{conclusion}. All 
the present data can be found at the Zenodo platform \cite{zenodo}.

@@

%=======================================================================
\section{Theoretical Method} 
\label{theory}
The energy loss of ions in metal targets responds to different physical 
mechanisms, depending on the impact ion velocity. At low velocities, the 
binary collisions are responsible for the loss of energy by the ion. 
The main contribution is the ionization of electrons of the metal 
conduction band, which is well approximated by a free electron gas (FEG) 
of Fermi velocity $v_F$. Above certain velocity (i.e. $v\geq 1.5\,v_F$), 
not only binary but also collective excitations (plasmons) 
\ale{occur}~\cite{mon17}. Moreover, at high energies, also the bound electrons 
contribute to the stopping power. The method used in this work combines 
a description for the interaction with the valence (or conduction) 
electrons as a FEG, and a different one for the interaction with the 
bound electrons.

@@

Hafnium ($Z=72$, [Xe] $4f^{14}\,6s^2\,5d_{3/2}^1\,5d_{5/2}^1$), belongs 
to the first groups of transition metals, with \ale{four} electrons as FEG 
(\st{theoretical} $r_S=2.14$ a.u.) and $1s$-$4f$ electrons bound. We compared 
the computed $r_S$ with the experimental value obtained from measured 
energy loss function by Lynch \textit{et al} \cite{lynch75}. The 
experimental plasmon energy of Hf is $\hbar\omega_P \approx 15.8$ eV, 
with a width at half maximum $\delta \approx 4.4$ eV, and 
$r_S \approx 2.07$ a.u. \cite{lynch75}. The difference of less than 
$5\%$ between theoretical and experimental $r_S$ assess Hf as a 
canonical target~\cite{mon17}.

@@

%----------------------------------------------------------------------
\begin{figure*}[!t]
\centering
\includegraphics[width=11.cm]{bindener.eps}
\caption{(a) Binding energies of Hf. \clau{Present relativistic and available non-relativistic~\cite{badnell97}  
values} are given with filled symbols. Experimental measurements 
for solids~\cite{williams1995} are depicted with open circles. 
(b) Corresponding relative errors respect to experimental data.}
\label{Binding_E}
\end{figure*}
%------------------------------------------------------------------------

For the stopping power due to bound electrons, the SLPA~\cite{mon17,mon13} 
is employed. It is worth \ale{mentioning} that the only inputs for the SLPA 
are the space-dependent densities of each shell in the ground state, 
and their binding energies. Collective processes and screening among 
electrons are included. \ale{Since} hafnium is a relativistic target, the 
wave functions and binding energies must be obtained by solving the 
\ale{many-electron} Dirac Hamiltonian. Details of these calculations 
\clau{and a table of binding energies} have been published in 
Ref.~\cite{mendez2019}, while Slater-type orbital expansions are 
given in \cite{Hf_arxiv}.

\ale{To} assess the importance of  \clau{a full relativistic} description of 
bound electrons, Fig.~\ref{Binding_E} (a) shows our \st{relativistic} binding 
energies, $E_{nl\pm}$, with $\pm=j\pm1/2$; the 
non-relativistic values~\cite{badnell97}; \clau{and the experimental data on 
solid Hf~\cite{williams1995}, which is available only for $1s$ to $4f_{\pm}$ 
subshells, as expected.} 
We notice that not only the most inner shells require relativistic calculations, but also the outer $5p$ and $4f$ shells. Furthermore, this figure shows very clearly the disability of non-relativistic calculations to describe the experimental data, which surprisingly worsens from the inner to the outer shells.

\clau{In the comparison with the experimental values in Fig.~\ref{Binding_E} (a) it can be noted that}
\ale{\st{Also,} the sign of the binding energy deviations is inverted}  \clau{for the outer $5s$-$4f\pm$ electrons, with the experimental binding energies being less bounded than our theoretical ones. Small differences for the outer shells are expected since } 
\ale{\st{can be attributed to the fact that} the experimental
values correspond to hafnium in solid-state, while our theoretical  calculations correspond to the element in the gas phase.}
%i.e. the electrons go from being more bounded --than the experimental values-- to being less bounded.

\clau{For more detail, Fig.~\ref{Binding_E} (b) shows}  the relative errors %\clau{of theoretical (relativistic and non relativistic)} 
with respect to the experimental  values. %\ale{\st{The relativistic binding energies of the electrons belonging to the deep K, L and M shells deviate less than $2.5\%$ from the experimental values, while the non-relativistic ones are within $\sim10\%$. Moreover, we notice that the relativistic corrections greatly influences the outer shells. This effect may be consequence of the orthogonality with the relativistic inner shells.}}
%\clau{ \st{This figure shows clearly that the relativistic corrections are decisive to describe the outer shells. The solution of the Dirac equation from inner to outer shells, and the orthogonality of the wave functions, involves all the shells.  }} 
\duda{This figure shows clearly that \st{for this target,} the relativistic corrections are very important to describe
the atomic structure of hafnium, even for the outer shells.
It turns out that the errors committed in the non-relativistic calculations of the inner shell orbitals 
propagate, through the Hartree-Fock approximation, to the outer shells.} \clau{The importance of full-relativistic calculations for the outer shells has already been noted for Au, Pb and Bi and W~\cite{mon09}.}

\clau{For the contribution of bound electrons to the total stopping cross sections}, the SLPA considers independent contributions of each subshell. Our relativistic binding energies present spin-orbit split. However, in total stopping power, where the initial state of the excited electron is not measured, the quantum uncertainty in energy $\Delta E$ melts this split. The criteria $\Delta E\Delta t\geq\hbar/2$ merges the energies 
$E_{nl+}-E_{nl-}$ for sufficiently small values of $\Delta t$ (the 
collisional mean-time). In fact, at sufficiently high impact velocity, 
we can expect all target electrons to respond together to the ion 
passage~\cite{lindhard53,chu72}. Following previous works~\cite{mon09},
the collisional time is estimated as $\Delta t\approx\langle r_i\rangle/v$, 
with $\langle r_i\rangle$ and $v$ being the orbital mean radio and 
impact velocity, respectively. 
\clau{In the case of hafnium}, we found that for every sub-shell of electrons, the spin-orbit split is unresolved in the energy region this sub-shell contributes. Therefore, the $nl$-electrons should be 
considered together, responding to the ion passage as a single gas of 
electrons with density $\delta_{nl}(r)$ and a mean binding energy 
$E_{nl}$. This feature is important within the SLPA calculations because 
it accounts for the screening among electrons of the \textit{same} 
binding energy. For example, the $4f_{-}$ and $4f_{+}$ of Hf can only 
be resolved for impact energies $E<0.05$ keV, but the contribution of 
$4f$ to the total stopping is negligible for $E<40$ keV. Moreover, the 
$5p$ and $4f$ electrons of Hf are very close in energy 
($\Delta E_{5p-4f} \approx 1$ a.u.~\cite{mendez2019}) and they react 
together for impact energies $E>40$ keV (inter-shell screening). \clau{As already mentioned, at higher energies inter-shell screening is possible for other subshells (i.e. 4p-4d for impact energies above 0.9 MeV), but its weight in the total stopping is minor for deeper shells.}

%-----------------------------------------------------------------------
\begin{figure*}[!t]
\centering
\includegraphics[width=13.cm]{Fig02_new3.eps}
\caption{(colour online) Theoretical stopping cross sections of protons in hafnium. \clau{Blue dash-dotted-line,the non-perturbative SPCC for the FEG; blue solid-line, ML results for the FEG (includes plasmon excitation); red
solid and dotted-lines, the SLPA results for bound electrons with and without $5p$-$4f$ screening, respectively. Black curves, total  stopping
adding the FEG and bound $1s$-$4f$ contributions: dash-dotted-line, SPCC (FEG) + SLPA (bound); solid-line, ML (FEG) + SLPA (bound) with 4f-5p screening; dotted-line, ML (FEG) + SLPA (bound) without 4f-5p screening.}
The vertical grey dashed-line indicates the energy \clau{of 37 keV} above which plasmon 
excitation is possible.} 
\label{slpa4f}
\end{figure*}
%-----------------------------------------------------------------------

\clau{Finally, in all our calculations~\cite{mon17} we assumed the projectile to be proton and not neutral hydrogen. When an ion moves inside a metal, the FEG screens the nucleus, so the binding energies will be smaller than outside the metal, and this effect is more important a low impact velocities $v$. In the case of hydrogen the difference is drastic, i.e. for H inside Hf ($rs=2.07$), the $1s$-bound state is almost null at $v<2$~\cite{suppression}. It is worth to mention that this assumption agrees with Ziegler SRIM code~\cite{Ziegler01}, but differs from CasP code~\cite{Grande} that predicts neutral hydrogen at very low velocities \st{based on measurements of charge states outside the metal}.}

%-----------------------------------------------------------------------
\begin{figure*}[!t]
\centering
\includegraphics[width=13.0cm]{Fig03_new2.eps}
\caption{(colour online) Stopping power cross section of hafnium for protons. Symbols: 
solid circles, present values; open circles, previous data~\cite{Sirotinin}. 
Curves: Black solid-line, present full theoretical results with 
$4f$-$5p$ screening; \clau{pink dash-dot line, theoretical CasP5.2~\cite{Grande,casp52} values; orange dash-double-dot line, theoretical DPASS~\cite{DPASS20} results; } green dotted-line, semi-empirical SRIM-2013~\cite{Ziegler01}; violet dashed-line, ICRU49~\cite{ICRU49} tabulated values.}
\label{F03}
\end{figure*}
%-----------------------------------------------------------------------

In Fig.~\ref{slpa4f}, we display the present theoretical stopping cross
section of Hf for protons \clau{using the relativistic wave functions and binding energies, but} with and without the $5p$-$4f$ screening. \clau{We show separately the FEG and bound electron contributions and the total stopping as the addition of both of them.}
\st{added the FEG and the bound electron contributions as explained above. }
The minimum energy for plasmon excitation was estimated $\sim 37$ keV. We
used the non-perturbative SPCC model for impact energies $E \leq 37$~keV, 
and the perturbative ML calculation above this energy. \clau{Bound $1s$-$4f$ electrons (relativistic wave functions and binding energies) are calculated with the SLPA and shown separately in Fig.~\ref{slpa4f} with and without the $4f$-$5p$ screening. Below $\sim 40$ keV the difference between both calculations is negligible}. Clearly, 
considering $5p$-$4f$ electrons as a single group of 20 electrons with 
screening among them gives lower stopping values than the addition of 
the separate $5p$ and $4f$ contributions. Notice that this shell 
correction can \st{only} be considered \duda{self-consistently? naturally? intrinsically?} within a many electron model such as the SLPA, .

@@

The present data \st{and theoretical results are} is displayed in 
Table~\ref{table01}. %, and Fig.~\ref{F03}. 
As can be \clau{observed,} %seen in Table~\ref{table01},
an overall relative uncertainty of around 5\% was 
achieved for the experimental stopping power values, which are mainly 
due to the uncertainty in the hafnium foil thickness.

@@

Fig.~\ref{F03} \clau{synthesizes the results of the present work.} \st{, we have good} \clau{The} agreement between the present 
theoretical results and the \clau{new} measurements displayed in Table~\ref{table01} \clau{is clear}. \clau{Present measurements using transmission method are in good agreement with the previous data by 
Sirotinin~\cite{Sirotinin}, measured in backscattering geometry}. Our theoretical approach also agrees with the data by 
Sirotinin~\cite{Sirotinin}, except for the lowest energy measurement at 
80 keV. \clau{We have also included in this figure the theoretical curves from Grande and Schiwietz CasP5.2 code~\cite{Grande,casp52} and from Sigmund and Schinner DPASS code~\cite{DPASS20}, both available online; also the semi-empirical results from SRIM-2013~\cite{Ziegler01} and the ICRU49 tables~\cite{ICRU49}}. It is interesting that our full theoretical curve differs from 
SRIM-2013 for impact energies below 100 keV. We obtain a stopping 
maximum around $40 \times 10^{-15}$ eV cm$^2$/atom  at 65 keV. Instead,
following the up-to-now only set of data \cite{Sirotinin}, SRIM-2013
suggests a lower stopping maximum  at impact energy of 115 kev. 
\clau{The stopping maximum is a very sensitive region for any full theoretical description, with no parameters at all, and this is quite visible in a linear-scale plot like Fig.~\ref{F03}. However, the impact energy for the maximum seem to agree between our curve and DPASS, although is $10 \%$ below. Instead, CasP predicts a stopping maximum $10 \%$ above our value and a completely different shape at lower energies.}
It is 
worth mentioning that \clau{our model gives} similar \st{theoretical} results using the experimental 
$r_S=2.07$ a.u. instead of the theoretical $r_S=2.14$ a.u. \st{give} \clau{with} the 
stopping maximum at the same impact energy but $4\%$ higher.

Future experiments \st{for impact energies below 100 keV} would be important to
complete this study \clau{on the stopping of Hf for protons. In the energy region $80-500$ keV only the measurements by Sirotinin group~\cite{Sirotinin} in 1984 are available. SRIM-2013 clearly includes these values in the adjust of the code, so it does not represent a test for this data. Undoubtedly, new measurements are needed in three specific energy regions: below 500 keV, around the stopping maximum (i.e. $50-200$ keV) and at low impact energies. }

@@

At high impact energies, the new stopping  measurements are in good 
agreement with our theoretical results, with previous experimental data
and semi-empirical values by SRIM-2013 and ICRU-49.  However, we call 
the attention that around the stopping maximum and at lower impact 
energies the difference between our full-theoretical results and SRIM 
is substantial. \clau{We compare our theoretical results with other two models given by the DPASS and CasP5.2 codes. Differences can be noted at intermediate to low impact energies, but  they also support an stopping maximum at lower energy than SRIM predictions}.

To the best of our knowledge, these are the first 
theoretical calculations of stopping in Hf taking into account the 
atomic relativistic effect \clau{and the screening among electrons in a consistent way from very low to high impact energies}. Future experiments for impact energies 
\clau{around the stopping maximum and in the low energy region} would be important to \st{complete  this study} \clau{have a more complete knowledge of the stopping of protons in hafnium}.

@@

\bibitem{Roth}
\ale{D. Roth, B. Bruckner, G. Undeutsch, V. Paneta, A. I. Mardare, 
C. L. McGahan, M. Dosmailov, J. I. Juaristi, M. Alducin, 
J. D. Pedarnig, R. F. Haglund, Jr., D. Primetzhofer, and P. Bauer
Phys. Rev. Lett. \textbf{119}, 163401 (2017).}

@@

\bibitem{Grande} \clau{G. Schiwietz, P. L. Grande, Nucl. Instrum. Meth. B \textbf{175-177}, 125 (2001); CasP code, available from www.casp-program.org}

\bibitem{casp52} \clau{G. Schiwietz and P. L. Grande,
Nucl. Instr. and Meth. B \textbf{273}, 1-5 (2012); P.L. Grande and G. Schiwietz,
Phys. Rev. A \textbf{58}, 3796 (1998).}

\bibitem{DPASS20} \clau{A. Schinner, P. Sigmund, Nucl. Instrum. Meth. B \textbf{460}, 19 (2019); P.Sigmund, A. Schinner, Eur. Phys. J. D \textbf{12}, 425 (2000). DPASS code, available from https://www.sdu.dk/en/DPASS/}

\bibitem{Ziegler01} 
J.F. Ziegler, J.P. Biersack, M. D. Ziegler, \clau{\textit{SRIM, The Stopping and Range of Ions in Matter}, (SRIM Co. Maryland, USA, 2008);} SRIM2013, 
Computer Program and Manual. Available from www.srim.org.

>>>>

%Title of paper
\title{The stopping power of hydrogen in hafnium and the importance of relativistic $4f$ electrons}
%Authors and affiliations

@@

%---------------------------Chile
\author{P. A. Miranda}
\affiliation{Departamento de F\'isica, Facultad de Ciencias Naturales, 
Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 
7800002, Chile}
%-------------------------Lisboa
\author{E. Alves}
\affiliation{Centro de Ci\^{e}ncias e Tecnologias Nucleares, Instituto 
Superior T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, 
Portugal}
\affiliation{Instituto de Plasma e Fus\~{a}o Nuclear, Instituto Superior 
T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, Portugal}

@@

So far, only one experimental work has been published regarding the 
stopping power cross section of pure hafnium for protons~\cite{Sirotinin}, 
while more attention has been recently given to  studies involving 
hafnium oxide due to its practical use~\cite{Abril,Behar,Primetzhofer,Roth}}. 
It is well known that significant attention has been paid in recent 
years to transition metal-oxides such as HfO$_2$ because of their 
potential as alternative gate dielectrics to replace SiO$_2$ for the 
future generation of nano-electronics with less than 45 nm gate 
length~\cite{Choi,Robertson}. Some important physical properties of the 
above mentioned metal-oxide films depend on its thickness, which is 
often measured by using Rutherford Backscattering Spectrometry~\cite{Alfassi01,Tesmer01}, 
a method that relies heavily on the determination of the stopping power 
of ion beams in the material of interest.

@@

The theoretical approach implemented in this work uses the shell-wise local plasma approximation (SLPA)~\cite{mon13} to describe the energy 
transferred to the bound $1s$-$4f$ electrons, and two different models 
for the FEG: in low energy region,
the screened potential with cusp condition model (SPCC)~\cite{mon17}, 
which is non-linear binary formalism; and the Mermin-Lindhard dielectric 
formalism (ML)~\cite{Mermin}, for energies around the stopping maximum 
and above. Our model requires the relativistic wave functions and 
binding energies of Hf, and considers 4 electrons per atom in the 
FEG~\cite{mendez2019}. Hf has the extra interest of the filled 
$4f$-subshell (with 14 electrons) as the main contribution below the 
FEG, causing the stopping cross sections to be very sensitive to a good 
description of this shell. The screening among the $4f$ and $5p$ 
electrons has been considered and found to play a major role within the 
SLPA calculations.

The experimental details and data are given in Section~\ref{experiment}, 
while the theoretical method is explained in Section~\ref{theory}. 
Present theoretical and experimental values are finally compared to the 
only experimental values by Sirotinin group~\cite{Sirotinin} measured 
with backscattering method, the theoretical results by  Grande and 
Schiwietz~\cite{Grande,casp52}, and by Sigmund and Schinner~\cite{DPASS20}, 
and also with the semi-empirical values from the SRIM-2013 
package~\cite{Ziegler01} and the ICRU-49 tabulation~\cite{ICRU49}}. 
Conclusions and discussions are given in Section~\ref{conclusion}. All 
the present data can be found at the Zenodo platform \cite{zenodo}.

@@

%=======================================================================
\section{Theoretical Method} 
\label{theory}
The energy loss of ions in metal targets responds to different physical 
mechanisms, depending on the impact ion velocity. At low velocities, the 
binary collisions are responsible for the loss of energy by the ion. 
The main contribution is the ionization of electrons of the metal 
conduction band, which is well approximated by a free electron gas (FEG) 
of Fermi velocity $v_F$. Above certain velocity (i.e. $v\geq 1.5\,v_F$), 
not only binary but also collective excitations (plasmons) 
occur}~\cite{mon17}. Moreover, at high energies, also the bound electrons 
contribute to the stopping power. The method used in this work combines 
a description for the interaction with the valence (or conduction) 
electrons as a FEG, and a different one for the interaction with the 
bound electrons.

@@

Hafnium ($Z=72$, [Xe] $4f^{14}\,6s^2\,5d_{3/2}^1\,5d_{5/2}^1$), belongs 
to the first groups of transition metals, with four} electrons as FEG 
($r_S=2.14$ a.u.) and $1s$-$4f$ electrons bound. We compared 
the computed $r_S$ with the experimental value obtained from measured 
energy loss function by Lynch \textit{et al} \cite{lynch75}. The 
experimental plasmon energy of Hf is $\hbar\omega_P \approx 15.8$ eV, 
with a width at half maximum $\delta \approx 4.4$ eV, and 
$r_S \approx 2.07$ a.u. \cite{lynch75}. The difference of less than 
$5\%$ between theoretical and experimental $r_S$ assess Hf as a 
canonical target~\cite{mon17}.

@@

%----------------------------------------------------------------------
\begin{figure*}[!t]
\centering
\includegraphics[width=11.cm]{bindener.eps}
\caption{(a) Binding energies of Hf. Present relativistic and available non-relativistic~\cite{badnell97}  
values are given with filled symbols. Experimental measurements 
for solids~\cite{williams1995} are depicted with open circles. 
(b) Corresponding relative errors respect to experimental data.}
\label{Binding_E}
\end{figure*}
%------------------------------------------------------------------------

For the stopping power due to bound electrons, the SLPA~\cite{mon17,mon13} 
is employed. It is worth mentioning} that the only inputs for the SLPA 
are the space-dependent densities of each shell in the ground state, 
and their binding energies. Collective processes and screening among 
electrons are included. Since} hafnium is a relativistic target, the 
wave functions and binding energies must be obtained by solving the 
many-electron} Dirac Hamiltonian. Details of these calculations 
and a table of binding energies have been published in 
Ref.~\cite{mendez2019}, while Slater-type orbital expansions are 
given in \cite{Hf_arxiv}.

To} assess the importance of a full relativistic description of 
bound electrons, Fig.~\ref{Binding_E} (a) shows our binding 
energies, $E_{nl\pm}$, with $\pm=j\pm1/2$; the 
non-relativistic values~\cite{badnell97}; and the experimental data on 
solid Hf~\cite{williams1995}, which is available only for $1s$ to $4f_{\pm}$ 
subshells, as expected.
We notice that not only the most inner shells require relativistic calculations, but also the outer $5p$ and $4f$ shells. Furthermore, this figure shows very clearly the disability of non-relativistic calculations to describe the experimental data, which surprisingly worsens from the inner to the outer shells.

In the comparison with the experimental values in Fig.~\ref{Binding_E} (a) 
it can be noted that the sign of the binding energy deviations is inverted 
for the outer $5s$-$4f\pm$ electrons, with the experimental binding energies 
being less bounded than our theoretical ones. Small differences for the 
outer shells are expected since the experimental values correspond to 
hafnium in solid-state, while our theoretical calculations correspond to 
the element in the gas phase.

For more detail, Fig.~\ref{Binding_E} (b) shows the relative errors 
with respect to the experimental  values. This figure shows clearly 
that the relativistic corrections are very 
important to describe the atomic structure of hafnium, even for the 
outer shells. It turns out that the errors committed in the 
non-relativistic calculations of the inner shell orbitals 
propagate, through the Hartree-Fock approximation, to the outer shells.
The importance of full-relativistic calculations for the outer 
shells has already been noted for Au, Pb and Bi and W~\cite{mon09}.

For the contribution of bound electrons to the total stopping 
cross sections, the SLPA considers independent contributions of each 
subshell. Our relativistic binding energies present spin-orbit split. 
However, in total stopping power, where the initial state of the 
excited electron is not measured, the quantum uncertainty in energy 
$\Delta E$ melts this split. The criteria $\Delta E\Delta t\geq\hbar/2$ 
merges the energies 
$E_{nl+}-E_{nl-}$ for sufficiently small values of $\Delta t$ (the 
collisional mean-time). In fact, at sufficiently high impact velocity, 
we can expect all target electrons to respond together to the ion 
passage~\cite{lindhard53,chu72}. Following previous works~\cite{mon09},
the collisional time is estimated as $\Delta t\approx\langle r_i\rangle/v$, 
with $\langle r_i\rangle$ and $v$ being the orbital mean radio and 
impact velocity, respectively. 
In the case of hafnium, we found that for every sub-shell of electrons, 
the spin-orbit split is unresolved in the energy region this sub-shell 
contributes. Therefore, the $nl$-electrons should be 
considered together, responding to the ion passage as a single gas of 
electrons with density $\delta_{nl}(r)$ and a mean binding energy 
$E_{nl}$. This feature is important within the SLPA calculations because 
it accounts for the screening among electrons of the \textit{same} 
binding energy. For example, the $4f_{-}$ and $4f_{+}$ of Hf can only 
be resolved for impact energies $E<0.05$ keV, but the contribution of 
$4f$ to the total stopping is negligible for $E<40$ keV. Moreover, the 
$5p$ and $4f$ electrons of Hf are very close in energy 
($\Delta E_{5p-4f} \approx 1$ a.u.~\cite{mendez2019}) and they react 
together for impact energies $E>40$ keV (inter-shell screening).
As already mentioned, at higher energies inter-shell screening is 
possible for other subshells (i.e. 4p-4d for impact energies above 
0.9 MeV), but its weight in the total stopping is minor for deeper shells.

%-----------------------------------------------------------------------
\begin{figure*}[!t]
\centering
\includegraphics[width=13.cm]{Fig02_new3.eps}
\caption{(colour online) Theoretical stopping cross sections of protons in hafnium. 
Blue dash-dotted-line,the non-perturbative SPCC for the FEG; blue solid-line, 
ML results for the FEG (includes plasmon excitation); red
solid and dotted-lines, the SLPA results for bound electrons with and 
without $5p$-$4f$ screening, respectively. Black curves, total  stopping
adding the FEG and bound $1s$-$4f$ contributions: dash-dotted-line, 
SPCC (FEG) + SLPA (bound); solid-line, ML (FEG) + SLPA (bound) with 
4f-5p screening; dotted-line, ML (FEG) + SLPA (bound) without 4f-5p screening.
The vertical grey dashed-line indicates the energy of 37 keV above which plasmon 
excitation is possible.} 
\label{slpa4f}
\end{figure*}
%-----------------------------------------------------------------------

Finally, in all our calculations~\cite{mon17} we assumed the projectile 
to be proton and not neutral hydrogen. When an ion moves inside a metal,
 the FEG screens the nucleus, so the binding energies will be smaller 
than outside the metal, and this effect is more important a low impact 
velocities $v$. In the case of hydrogen the difference is drastic, i.e. 
for H inside Hf ($rs=2.07$), the $1s$-bound state is almost null at 
$v<2$~\cite{suppression}. It is worth to mention that this assumption 
agrees with Ziegler SRIM code~\cite{Ziegler01}, but differs from CasP 
code~\cite{Grande} that predicts neutral hydrogen at very low velocities.

%-----------------------------------------------------------------------
\begin{figure*}[!t]
\centering
\includegraphics[width=13.0cm]{Fig03_new2.eps}
\caption{(colour online) Stopping power cross section of hafnium for protons. Symbols: 
solid circles, present values; open circles, previous data~\cite{Sirotinin}. 
Curves: Black solid-line, present full theoretical results with 
$4f$-$5p$ screening; pink dash-dot line, theoretical 
CasP5.2~\cite{Grande,casp52} values; orange dash-double-dot line, 
theoretical DPASS~\cite{DPASS20} results; green dotted-line, 
semi-empirical SRIM-2013~\cite{Ziegler01}; violet dashed-line, 
ICRU49~\cite{ICRU49} tabulated values.}
\label{F03}
\end{figure*}
%-----------------------------------------------------------------------

In Fig.~\ref{slpa4f}, we display the present theoretical stopping cross
section of Hf for protons using the relativistic wave functions and
 binding energies, but with and without the $5p$-$4f$ screening. 
We show separately the FEG and bound electron contributions and the 
total stopping as the addition of both of them.
The minimum energy for plasmon excitation was estimated $\sim 37$ keV. We
used the non-perturbative SPCC model for impact energies $E \leq 37$~keV, 
and the perturbative ML calculation above this energy. Bound $1s$-$4f$ 
electrons (relativistic wave functions and binding energies) are 
calculated with the SLPA and shown separately in Fig.~\ref{slpa4f} 
with and without the $4f$-$5p$ screening. Below $\sim 40$ keV the 
difference between both calculations is negligible. Clearly, 
considering $5p$-$4f$ electrons as a single group of 20 electrons with 
screening among them gives lower stopping values than the addition of 
the separate $5p$ and $4f$ contributions. Notice that this shell 
correction can be considered naturally within a many electron model 
such as the SLPA, .

@@

The present data is displayed in 
Table~\ref{table01}. %, and Fig.~\ref{F03}. 
As can be observed, %seen in Table~\ref{table01},
an overall relative uncertainty of around 5\% was 
achieved for the experimental stopping power values, which are mainly 
due to the uncertainty in the hafnium foil thickness.

@@

Fig.~\ref{F03} synthesizes the results of the present work.
The agreement between the present theoretical results and the new
measurements displayed in Table~\ref{table01} is clear. Present 
measurements using transmission method are in good agreement with the 
previous data by Sirotinin~\cite{Sirotinin}, measured in backscattering 
geometry. Our theoretical approach also agrees with the data by 
Sirotinin~\cite{Sirotinin}, except for the lowest energy measurement at 
80 keV. We have also included in this figure the theoretical curves from 
Grande and Schiwietz CasP5.2 code~\cite{Grande,casp52} and from Sigmund 
and Schinner DPASS code~\cite{DPASS20}, both available online; also the 
semi-empirical results from SRIM-2013~\cite{Ziegler01} and the ICRU49 
tables~\cite{ICRU49}. It is interesting that our full theoretical curve 
differs from SRIM-2013 for impact energies below 100 keV. We obtain a 
stopping maximum around $40 \times 10^{-15}$ eV cm$^2$/atom  at 65 keV. 
Instead, following the up-to-now only set of data \cite{Sirotinin}, SRIM-2013
suggests a lower stopping maximum  at impact energy of 115 kev. 
The stopping maximum is a very sensitive region for any full theoretical 
description, with no parameters at all, and this is quite visible in a 
linear-scale plot like Fig.~\ref{F03}. However, the impact energy for 
the maximum seem to agree between our curve and DPASS, although is 
$10 \%$ below. Instead, CasP predicts a stopping maximum $10 \%$ above 
our value and a completely different shape at lower energies. It is 
worth mentioning that our model gives similar results using the experimental 
$r_S=2.07$ a.u. instead of the theoretical $r_S=2.14$ a.u. with the 
stopping maximum at the same impact energy but $4\%$ higher.

Future experiments would be important to complete this study on the 
stopping of Hf for protons. In the energy region $80-500$ keV only 
the measurements by Sirotinin group~\cite{Sirotinin} in 1984 are 
available. SRIM-2013 clearly includes these values in the adjust of 
the code, so it does not represent a test for this data. Undoubtedly, 
new measurements are needed in three specific energy regions: 
below 500 keV, around the stopping maximum (i.e. $50-200$ keV) and at 
low impact energies.

@@

At high impact energies, the new stopping  measurements are in good 
agreement with our theoretical results, with previous experimental data
and semi-empirical values by SRIM-2013 and ICRU-49.  However, we call 
the attention that around the stopping maximum and at lower impact 
energies the difference between our full-theoretical results and SRIM 
is substantial. We compare our theoretical results with other two 
models given by the DPASS and CasP5.2 codes. Differences can be noted 
at intermediate to low impact energies, but  they also support an 
stopping maximum at lower energy than SRIM predictions.

To the best of our knowledge, these are the first 
theoretical calculations of stopping in Hf taking into account the 
atomic relativistic effect and the screening among electrons in a 
consistent way from very low to high impact energies. Future 
experiments for impact energies around the stopping maximum and in 
the low energy region would be important to have a more complete 
knowledge of the stopping of protons in hafnium.

@@

\bibitem{Roth}
D. Roth, B. Bruckner, G. Undeutsch, V. Paneta, A. I. Mardare, 
C. L. McGahan, M. Dosmailov, J. I. Juaristi, M. Alducin, 
J. D. Pedarnig, R. F. Haglund, Jr., D. Primetzhofer, and P. Bauer
Phys. Rev. Lett. \textbf{119}, 163401 (2017).

@@

\bibitem{Grande} 
G. Schiwietz, P. L. Grande, 
Nucl. Instrum. Meth. B \textbf{175-177}, 125 (2001); 
CasP code, available from www.casp-program.org

\bibitem{casp52} 
G. Schiwietz and P. L. Grande,
Nucl. Instr. and Meth. B \textbf{273}, 1-5 (2012); 
P.L. Grande and G. Schiwietz,
Phys. Rev. A \textbf{58}, 3796 (1998).

\bibitem{DPASS20}
A. Schinner, P. Sigmund, 
Nucl. Instrum. Meth. B \textbf{460}, 19 (2019); 
P.Sigmund, A. Schinner, 
Eur. Phys. J. D \textbf{12}, 425 (2000). 
DPASS code, available from https://www.sdu.dk/en/DPASS/

\bibitem{Ziegler01} 
J.F. Ziegler, J.P. Biersack, M. D. Ziegler, 
\textit{SRIM, The Stopping and Range of Ions in Matter}, 
(SRIM Co. Maryland, USA, 2008); 
SRIM2013, Computer Program and Manual. Available from www.srim.org.
